Astrophysical calibration of gravitational-wave detectors

### 2015-07-24 docs/cosmic_cal.tex
Added posterior plot showing example of NS-BH system, and added more text

For each of the simulated sources we have used the {\tt emcee} python MCMC package to perform 
parameter estimation over the unknown source parameters using the prior as discussed in 
Section~\ref{sec:priors}. In each case we use an estimate of the noise PSD based on the design 
sensitivity, but simulated as if estimated by averaging 64 separate PSD estimates and scaled with 
the same factor as applied to the injection and noise\footnote{We do not account for there being 
a difference between the estimated PSD and the actual PSD of the analysed section of data. This 
difference would be very much correlated with the scale factor, so in reality our estimate of the 
scale factor would be a combination of the calibration offset and the difference in the PSD. As 
such our results on real data would be an upper limit on the calibration scale factor.}. This has 
provided posterior probability distributions on the calibration scale factors for each detector 
\MP{Show an example for both source types}. An example of the marginalised posterior probablil From 
these we have calculated the 68\% credible region 
on the scale factors for each detector (if these were Gaussian distributions this is equivalent to 
the region bounded by $1\sigma$ intervals). For all the signals at each distance increment we have 
produced the distribution of the fractional half widths (i.e.\ $1\sigma$) of these scale factors 
confidence intervals compared to the actual value. These are shown as boxplots in 
figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} and \ac{NSBH} systems 
respectively. The boxes show the extent from the lower to upper quartile of the values, whilst the 
whiskers represent range of values. The black line within each box gives the median value and the 
star gives the mean value. Also shown on each plot as the dashed magenta line is the percentage of 
proposed injections that actually fulfilled the detection criterion.

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{bns_post_fig.pdf}
 \end{center}
 \caption{\label{fig:bnspost} The marginalised posterior probability distributions for the unknown
 parameters of a \ac{BNS} system including calibration scaling factors for the three detectors (H1, 
L1 and V1). The injected signal had \ac{SNR} of 8.7, 10.9 and 4.1 and percentage uncertainties in 
the calibration scale factors of 14\%, 11\% and 27\% for each of the detectors respectively.}
\end{figure}

@@

In figure~\ref{fig:bnsresults} we see that unsurprisingly on average the scale factor can be 
recovered to equivalent precision in both \ac{aLIGO} detectors, with uncertainties generally within 
the 10\% range for sources at 100\,Mpc. An interesting feature that we see is that for distances 
$\gtrsim 250$\,Mpc the upper extent of the uncertainty hits a maximum at $\sim 25\mbox{--}30\%$, 
whilst the width of boxes narrows. This is due to our ``detectability'' criteria, in that for all 
distances we will only be seeing those sources with high enough SNR that we would consider them 
detectable. There will therefore be fewer sources with SNR higher than this criteria at large 
distances, giving us a narrower range, and we automatically exclude those with SNR that are too 
small thus truncating our uncertainty distribution at the upper range. However, this does show that 
on average for sources that are detectable out to 450\,Mpc we would be able to constrain the 
calibration scale factor uncertainty to $\sim 20\%$. As the detection criteria will generally come 
from the SNR being high enough within the two \ac{aLIGO} instruments this means that the SNR in 
\ac{AdV} can be small and thus the ability to constrain the scale factor for it becomes poor 
(although it still provides information that the calibration is not grossly inaccurate).

In figure~\ref{fig:bnsresults} we see slightly different results. Despite the generally higher SNR 
for the \ac{NSBH} systems we find that on averge that the scale factor uncertainty is still roughly 
20\% at distances to 400--500\,Mpc. The range of uncertainty values is also far broader at all 
distances and there is not the plateau and upper limit seen for the \ac{BNS} systems. \MP{This 
needs checked, but ... This broader range of uncertainties appears to be down to the inclusion of 
the spin parameter and correlations this has with the mass ratio $q$ - show posterior examples. The 
lack of an upper plateau seems to be due to the detection criteria having a different effect on the 
distribution of sources that could be seen, in particular it biases the observable sources towards 
having a higher mass ratio (again check this) and thus having more correlation with the spin.}

@@

This paper has shown that for differing levels of SNR, the error in the
calibration is resolved within a certain accuracy range. Initial simulations
used a bad sky position, resulting a low SNR which created larger uncertainty
in resolving $s$. It was shown that for this sky position, the calibration
scale factor estimation was recovered with an accuracy of roughly a $27\%$ for
$D_{L} = 200\,Mpc$ for LIGO Hanford. The resolution of $s$ for a network of
detectors was also shown.

The main results obtained in this project were from analysing randomly chosen
sky positions from the entire field of view for the detector. Through these
simulations, it was shown that this new method of calibration can reach and
surpass the current method in terms of accuracy.

These results are reassuring for parameter estimation of the defining GW signal
and would help in reducing the uncertainty in a direct detection. This project
was; however, time constrained. There are many, many things that could be done
with this project. A true noise realisation with each frequency bin having a
random Gaussian noise would be the next step in calibration error estimation.

>>>>

A simple assumption that one could make would be that the uncertainties on the calibration scale 
factors (if they are independent of other parameters and have a Gaussian probability distribtuion) 
should be given by $\sim 1/{\rm SNR}$ for each detector. As we will show below this assumption is 
reasonable, but correlations do exist between parameters meaning that it does not completely hold, 
especially in the case on \ac{NSBH} systems with a spinning component.

For each of the simulated sources we have used the {\tt emcee} python MCMC package to perform 
parameter estimation over the unknown source parameters using the prior as discussed in 
Section~\ref{sec:priors}. In each case we use an estimate of the noise PSD based on the design 
sensitivity, but simulated as if estimated by averaging 64 separate PSD estimates and scaled with 
the same factor as applied to the injection and noise\footnote{We do not account for there being 
a difference between the estimated PSD and the actual PSD of the analysed section of data. This 
difference would be very much correlated with the scale factor, so in reality our estimate of the 
scale factor would be a combination of the calibration offset and the difference in the PSD. As 
such our results on real data would be an upper limit on the calibration scale factor.}. This has 
provided posterior probability distributions on the calibration scale factors for each detector 
\MP{Show an example for both source types}. Examples of the marginalised posterior probabilities
for a \ac{BNS} system and a \ac{NSBH} system observed with the three detector network are shown in 
figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

From these posterior distributions we have calculated the 68\% credible region on the scale factors 
for each detector (if these were Gaussian distributions this is equivalent to the region bounded by 
$1\sigma$ intervals). For all the signals at each distance increment we have 
produced the distribution of the fractional half widths (i.e.\ $1\sigma$) of these scale factors 
confidence intervals compared to the actual value. These are shown as boxplots in 
figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} and \ac{NSBH} systems 
respectively. The boxes show the extent from the lower to upper quartile of the values, whilst the 
whiskers represent the range of values. The black line within each box gives the median value and 
the star gives the mean value. Also shown on each plot as the dashed magenta line is the percentage 
of proposed injections that actually fulfilled the detection criterion.

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{bns_post_fig.pdf}
 \end{center}
 \caption{\label{fig:bnspost} The marginalised posterior probability distributions for the unknown
 parameters of a \ac{BNS} system including calibration scaling factors for the three detectors (H1, 
L1 and V1). The injected signal was at a distance of 250\,Mpc, had \ac{SNR} of 8.7, 10.9 and 4.1 
and percentage uncertainties in the calibration scale factors of 14\%, 11\% and 27\% for each of 
the detectors respectively.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{nsbh_post_fig.pdf}
 \end{center}
 \caption{\label{fig:nsbhpost} The marginalised posterior probability distributions for the unknown
 parameters of a \ac{NSBH} system including calibration scaling factors for the three detectors 
(H1, L1 and V1). The injected signal was at a distance of 450\,Mpc, had \ac{SNR} of 7.7, 10.9 and 
4.2 and percentage uncertainties in the calibration scale factors of 17\%, 14\% and 34\% for each of 
the detectors respectively.}
\end{figure}

@@

In figure~\ref{fig:bnsresults} we see that unsurprisingly on average the scale factor can be 
recovered to equivalent precision in both \ac{aLIGO} detectors, with uncertainties generally within 
the 10\% range for sources at 100\,Mpc. An interesting feature that we see is that for distances 
$\gtrsim 250$\,Mpc the upper extent of the uncertainty hits a maximum at $\sim 25\mbox{--}30\%$, 
whilst the width of boxes narrows. This is due to our ``detectability'' criteria, in that for all 
distances we will only be seeing those sources with high enough SNR that we would consider them 
detectable. There will therefore be fewer sources with SNR higher than this criteria at large 
distances, giving us a narrower range, and we automatically exclude those with SNR that are too 
small thus truncating our uncertainty distribution at the upper range. However, this does show that 
on average for sources that are detectable out to 450\,Mpc we would be able to constrain the 
calibration scale factor uncertainty to $\sim 20\%$. As the detection criteria will generally come 
from the SNR being high enough within the two \ac{aLIGO} instruments this means that the SNR in 
\ac{AdV} can be small and thus the ability to constrain the scale factor for it becomes poor 
(although it still provides information that the calibration is not grossly inaccurate). We also 
see that true uncertainties achievable for recovering the calibration scale factors are indeed very 
similar to the simple assumption that they would be given by $sim 1/{\rm SNR}$. Although, they are 
probably slightly poorer due to their still be correlations with $\iota$ even within the 
constrained prior range, and slight correlation between the scale factors for H1 and L1 as seen in 
figure~\ref{fig:bnspost}.

In figure~\ref{fig:nsbhresults} we see slightly different results. Despite the generally higher SNR 
for the \ac{NSBH} systems we find that on averge that the scale factor uncertainty is still roughly 
20\% at distances to 400--500\,Mpc. The range of uncertainty values is also far broader at all 
distances and there is not the plateau and upper limit seen for the \ac{BNS} systems. \MP{This 
needs checked, but ... This broader range of uncertainties appears to be down to the inclusion of 
the spin parameter and correlations this has with the mass ratio $q$ - show posterior examples. The 
lack of an upper plateau seems to be due to the detection criteria having a different effect on the 
distribution of sources that could be seen, in particular it biases the observable sources towards 
having a higher mass ratio (again check this) and thus having more correlation with the spin.}

@@
